---
title: "Final Project"
author: "Group 2"
output:
  html_document:
    css: style_1.css
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE,message=FALSE)
```
#### Load Dataset and check for empty values

```{r}
# Load dataset
data <- read.csv('Dataset.csv')
sum(is.na(data))
```

#### Check for duplicate rows

```{r}
# Check for duplicate rows
duplicate_rows <- sum(duplicated(data))
duplicate_rows
```

#### Total number of rows

```{r}
sum(complete.cases(data))
```

#### Columns in Dataset

```{r}
library(dplyr)
# Rename the columns
data <- data %>%
  rename(
    Patient_Number = No_Pation,
    Creatinine = Cr,
    HbA1c_Level = HbA1c,
    Cholesterol = Chol,
    Triglycerides = TG,
    HDL_Cholesterol = HDL,
    LDL_Cholesterol = LDL,
    VLDL_Cholesterol = VLDL
  )

colnames(data)
```

#### Removing leading and trailing spaces

Before removing the spaces

```{r}

table(data$CLASS)

# Apply trimws to all columns in the dataset
data <- data.frame(lapply(data, function(x) {
  if (is.character(x)) {
    trimws(x) # Trim whitespace for character columns
  } else {
    x # Leave other columns unchanged
  }
}), stringsAsFactors = FALSE)
```

After removing the spaces

```{r}
table(data$CLASS)

```

#### Calculate summary statistics (mean, median, standard deviation) for BMI and lipid profiles (LDL, HDL, TG), stratified by CLASS (diabetic vs. non-diabetic).

```{r}
library(knitr)
library(kableExtra)

# Assuming `data` is your dataframe loaded into R
# Group by CLASS and calculate summary statistics
summary_stats <- data %>%
  group_by(CLASS) %>%
  summarise(
    BMI_mean = mean(BMI, na.rm = TRUE),
    BMI_median = median(BMI, na.rm = TRUE),
    BMI_sd = sd(BMI, na.rm = TRUE),
    LDL_mean = mean(LDL_Cholesterol, na.rm = TRUE),
    LDL_median = median(LDL_Cholesterol, na.rm = TRUE),
    LDL_sd = sd(LDL_Cholesterol, na.rm = TRUE),
    HDL_mean = mean(HDL_Cholesterol, na.rm = TRUE),
    HDL_median = median(HDL_Cholesterol, na.rm = TRUE),
    HDL_sd = sd(HDL_Cholesterol, na.rm = TRUE),
    TG_mean = mean(Triglycerides, na.rm = TRUE),
    TG_median = median(Triglycerides, na.rm = TRUE),
    TG_sd = sd(Triglycerides, na.rm = TRUE)
  )

# View the results
summary_stats %>%
  kable(format = "html", digits = 2) %>%
  kable_styling(font_size = 10)  # Adjust the font size as needed
```

#### Machine learning models for predicting the class of diabetes (N, P, Y) based on clinical and demographic factors?

### Random Forest

```{r, echo= FALSE}
# Load necessary library
library(dplyr)
library(ggplot2)
library(psych)
library(caret)
library(randomForest)
library(class)
library(rpart)
library(rpart.plot)
library(partykit)
library(corrplot)
library(gridExtra)
library(rpart)
library(rattle)
```

```{r}
data$CLASS <- as.factor(data$CLASS)

set.seed(42)  # For reproducibility
# Split data -> 80% as training data and 20% as testing data
train_index <- createDataPartition(data$CLASS, p = 0.8, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]

# Fit a Random Forest model
rf_model <- randomForest(CLASS ~ ., data = train_data, ntree = 100, importance = TRUE,random_state = 42)

# Make predictions on the test set
predictions <- predict(rf_model, test_data)

# Evaluate the model
conf_matrix <- confusionMatrix(predictions, test_data$CLASS)
print(conf_matrix)
```

```{r}
# Get variable importance from the fitted model
var_importance <- randomForest::importance(rf_model)

# Print variable importance
print(var_importance)
# Plot variable importance
varImpPlot(rf_model, main = "Variable Importance")
```
-----------------------------------------------------------------------------------------------------------------

### K-Nearest Neighbour

```{r}
data$CLASS <- as.factor(data$CLASS)  # Ensure CLASS is a factor

# Scale numeric features
numeric_columns <- c("AGE", "Urea", "Creatinine", "HbA1c_Level", "Cholesterol", "BMI")
train_scaled <- scale(train_data[, numeric_columns])
test_scaled <- scale(test_data[, numeric_columns])

# Cross-validation to find the optimal k
set.seed(42)
error<-rep(NA,20) # Placeholder

for (i in 1:20) {
  # Perform KNN
  knn_pred <- knn(train = train_scaled, test = test_scaled, cl = train_data$CLASS, k = i)
  
  # Calculate test error
  error[i] <- mean(knn_pred != test_data$CLASS)
}

error_df <- data.frame(
  K = 1:20,                   # Number of neighbors
  Error = error               # Test error rates
)

# Add Accuracy (1 - Error) to the data frame
error_df$Accuracy <- 1 - error_df$Error

# Plot accuracy vs. K using ggplot2
ggplot(error_df, aes(x = K, y = Accuracy)) +
  geom_line(color = "blue") +
  geom_point() +
  ggtitle("Accuracy vs K for KNN") +
  xlab("Number of Neighbors (K)") +
  ylab("Accuracy") +
  theme_minimal()

ggplot(error_df, aes(x = K, y = Error)) +
  geom_line(color = "blue") +
  geom_point() +
  ggtitle("Error vs K for KNN") +
  xlab("Number of Neighbors (K)") +
  ylab("Error") +
  theme_minimal()

# Find the minimum error and corresponding K value
min_error <- min(error_df$Error)
optimal_k <- error_df$K[which.min(error_df$Error)]

# Print the results
print(paste("Minimum Error:", round(min_error, 4)))
print(paste("Optimal K:", optimal_k))

```

-----------------------------------------------------------------------------------------------------------------

### Decision Tree

```{r}
###Decision Tree

# Fit a Decision Tree model
dt_model <- rpart(CLASS ~ ., data = train_data, method = "class")

# Plot the Decision Tree
fancyRpartPlot(dt_model)

# Make predictions on the test data
predictions <- predict(dt_model, test_data, type = "class")

# Evaluate the model
conf_matrix <- confusionMatrix(predictions, test_data$CLASS)
print(conf_matrix)

# Optional: Print overall accuracy
accuracy <- conf_matrix$overall["Accuracy"]
cat("Accuracy:", accuracy, "\n")

```


-----------------------------------------------------------------------------------------------------------------

#### Use of BMI thresholds to classify individuals into non-diabetic, pre-diabetic, and diabetic categories, and the role BMI plays in predicting diabetes progression across these classes

```{r}
# Density plot to compare BMI distribution
ggplot(data, aes(x = BMI, fill = CLASS)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(breaks=seq(0,48,by=2))+
  labs(title = "BMI Density Plot by CLASS", x = "BMI", y = "Density") +
  theme_minimal()
```


-----------------------------------------------------------------------------------------------------------------

#### Gender-specific differences in clinical markers or diabetes class distributions?

```{r}
# Correlation Heatmaps by Gender with Values in Boxes
# Subset data by gender
data_male <- subset(data, Gender == "M")
data_female <- subset(data, Gender == "F")
  
# Calculate correlation matrix
correlation_matrix_male <- cor(data_male[, c("HbA1c_Level", "BMI", "HDL_Cholesterol", "LDL_Cholesterol", "Triglycerides", "AGE", "Urea", "Creatinine")], use = "complete.obs")

correlation_matrix_female <- cor(data_female[, c("HbA1c_Level", "BMI", "HDL_Cholesterol", "LDL_Cholesterol", "Triglycerides", "AGE", "Urea", "Creatinine")], use = "complete.obs")
  
# Plot heatmap with values
corrplot(correlation_matrix_male, method = "color", 
           col = colorRampPalette(c("blue", "white", "red"))(200), 
           addCoef.col = "black", # Add values to the boxes in black
           tl.col = "black",      # Labels in black
           tl.cex = 0.8,          # Adjust label size
           number.cex = 0.7,      # Adjust coefficient size
           title = paste("Correlation Heatmap for Gender: M"), mar = c(0, 0, 1, 0))

corrplot(correlation_matrix_female, method = "color", 
           col = colorRampPalette(c("blue", "white", "red"))(200), 
           addCoef.col = "black", # Add values to the boxes in black
           tl.col = "black",      # Labels in black
           tl.cex = 0.8,          # Adjust label size
           number.cex = 0.7,      # Adjust coefficient size
           title = paste("Correlation Heatmap for Gender: F"), mar = c(0, 0, 1, 0))

#linear model
lm_female <- lm(Creatinine~Urea, data_female)
lm_male <- lm(Creatinine~Urea, data_male)


# Scatterplot for males
plot_male <- ggplot(data_male, aes(x = Urea, y = Creatinine)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(intercept = coef(lm_male)[1], slope = coef(lm_male)[2], color = "black") +
  labs(title = "Scatterplot of Urea vs Creatine (Males)",
       x = "Urea", y = "Creatine") +
  theme_minimal()

# Scatterplot for females
plot_female <- ggplot(data_female, aes(x = Urea, y = Creatinine)) +
  geom_point(color = "red", alpha = 0.7) +
  geom_abline(intercept = coef(lm_female)[1], slope = coef(lm_female)[2], color = "black") +
  labs(title = "Scatterplot of Urea vs Creatine (Females)",
       x = "Urea", y = "Creatine") +
  theme_minimal()

# Arrange the plots side by side
grid.arrange(plot_male, plot_female, ncol = 2)
##########

#optimize

# Fit a polynomial regression model (degree 2)
poly_model <- lm(Creatinine ~ poly(Urea, 2), data = data_female)

# Make predictions
data_female$Cr_pred <- predict(poly_model, newdata = data_female)

# Evaluate the model: Mean Squared Error and R-squared
mse_poly <- mean((data_female$Creatinine - data_female$Cr_pred)^2)
r2_poly <- summary(poly_model)$r.squared

# Visualization
ggplot(data_female, aes(x = Urea, y = Creatinine)) +
  geom_point(alpha = 0.7, color = "purple", label = "Data Points") +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2), color = "blue", se = FALSE, label = "Polynomial Regression Curve") +
  labs(title = "Optimized Polynomial Regression: Urea vs. Creatinine (Gender: F)",
       x = "Urea", y = "Creatinine") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_line(aes(y = Cr_pred), color = "blue")

# Print evaluation metrics
cat("Evaluation metrics\n",
    "Mean Squared Error (MSE):", mse_poly, "\n",
    "R-squared (R²):",r2_poly,"\n")

# Fitting the polynomial model
poly_model <- lm(Creatinine ~ poly(Urea, 2), data = data_female)

# Predictions
predictions <- predict(poly_model, newdata = data_female)

# Residuals
residuals <- data_female$Creatinine - predictions

# Metrics
mae <- mean(abs(residuals))
mape <- mean(abs(residuals / data_female$Creatinine)) * 100
accuracy <- 100 - mape
r_squared <- summary(poly_model)$r.squared

# Output the results
cat("Results for Females\n",
    "---------------------\n",
    "Mean Absolute Error (MAE):", mae, "\n",
    "Mean Absolute Percentage Error (MAPE):", mape, "%\n",
    "Model Accuracy:", accuracy, "%\n",
    "R-squared (R²):", r_squared, "\n")

# Fitting the polynomial model for males
poly_model_male <- lm(Creatinine ~ poly(Urea, 2), data = data_male)

# Predictions for males
predictions_male <- predict(poly_model_male, newdata = data_male)

# Residuals for males
residuals_male <- data_male$Creatinine - predictions_male

# Metrics for males
mae_male <- mean(abs(residuals_male))  # Mean Absolute Error
mape_male <- mean(abs(residuals_male / data_male$Creatinine)) * 100  # Mean Absolute Percentage Error
accuracy_male <- 100 - mape_male  # Accuracy
r_squared_male <- summary(poly_model_male)$r.squared  # R-squared

# Output the results for males
cat("Results for Males\n",
    "---------------------\n",
    "Mean Absolute Error (MAE) for Males:", mae_male, "\n",
    "Mean Absolute Percentage Error (MAPE) for Males:", mape_male, "%\n",
    "Model Accuracy for Males:", accuracy_male, "%\n",
    "R-squared (R²) for Males:", r_squared_male, "\n")




```

